---
title: "team-navy-nyc-delays-lab"
format: html
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(nycflights13)
library(ggrepel)
## Inline data preparation (reads carriers.csv and prepares `departures`)

# ensure data directory
if (!dir.exists("data")) {
  dir.create("data")
}

# read carriers lookup
carriers <- readr::read_csv("carriers.csv", show_col_types = FALSE)

# prepare departures tibble per lab instructions
departures <- flights %>%
  filter(!is.na(dep_delay)) %>%
  transmute(
    airline_code = carrier,
    airport_code = origin,
    airport = dplyr::recode(
      origin,
      EWR = "Newark",
      LGA = "LaGuardia",
      JFK = "JFK"
    ),
    hour = hour,
    on_time = dep_delay < 15
  ) %>%
  left_join(carriers, by = c("airline_code" = "airline_code")) %>%
  select(airline = airline, airport, hour, on_time)

# save prepared data for reproducibility (optional)
readr::write_rds(departures, file = "data/departures.rds")
```

## 1 Overview

This Quarto document implements the Week 06 lab tasks on visualizing on-time
departures from New York City-area airports. Follow the code chunks below to
reproduce the figures and answers from the lab brief.

## 2 Data preparation

The data-preparation code is in `scripts/prepare_departures.R` and saves a
prepared `data/departures.rds` file.

## 3 Tasks

Placeholders for lab tasks (3.1–3.9). Implement plots in subsequent code
chunks following the lab instructions.

<!-- data is prepared in the setup chunk and available as `departures` -->

### 3.1 Data preparation checks

```{r}
#| label: prep-checks
# Verifier: ensure columns exist and no missing values
stopifnot(all(
  c("airline", "airport", "hour", "on_time") %in% names(departures)
))
stopifnot(!any(is.na(departures)))
```

### 3.2 Airlines Ranked by Departures (Lollipop)

```{r}
#| label: departures-by-airline
departures_by_airline <- departures %>%
  group_by(airline) %>%
  summarise(n = n(), on_time_pct = mean(on_time) * 100, .groups = "drop") %>%
  arrange(n)

# convert to factor ordered by n
departures_by_airline <- departures_by_airline %>%
  mutate(airline = factor(airline, levels = airline))

library(ggplot2)
library(forcats)

ggplot(departures_by_airline, aes(x = n, y = fct_reorder(airline, n))) +
  geom_segment(
    aes(x = 0, xend = n, y = airline, yend = airline),
    color = "gray70"
  ) +
  geom_point(size = 3, color = "steelblue") +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  labs(
    x = "Number of departures (n)",
    y = "Airline",
    title = "Airlines Ranked by Number of Departures"
  ) +
  theme_minimal() +
  theme(axis.title.y = element_blank()) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)))
```

```{r}
# Verifier: sum of counts equals number of rows in departures
stopifnot(sum(departures_by_airline$n) == nrow(departures))
```

### 3.3 On-Time Departures by Airline (Cleveland dot plot)

```{r}
#| label: punctuality-ranking
# remove airlines with fewer than 1000 departures
departures_by_airline_big <- departures_by_airline %>% filter(n >= 1000)

departures_by_airline_big <- departures_by_airline_big %>%
  arrange(on_time_pct) %>%
  mutate(airline = factor(airline, levels = airline))

ggplot(departures_by_airline_big, aes(x = on_time_pct, y = fct_rev(airline))) +
  geom_point(size = 3, color = "darkgreen") +
  geom_text(
    aes(label = sprintf("%.1f%%", on_time_pct)),
    hjust = -0.1,
    size = 3
  ) +
  labs(
    x = "On-time departure percentage",
    y = "Airline",
    title = "Airlines Ranked by On-time Percentage (>=1000 departures)"
  ) +
  theme_minimal() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05)))
```

```{r}
# Verifier: check busiest airline's on-time pct matches calculation from departures
busiest_airline <- departures_by_airline %>%
  arrange(desc(n)) %>%
  slice_head(n = 1) %>%
  pull(airline)
calc_pct <- departures %>%
  filter(airline == busiest_airline) %>%
  summarise(pct = mean(on_time) * 100) %>%
  pull(pct)
found_pct <- departures_by_airline %>%
  filter(airline == busiest_airline) %>%
  pull(on_time_pct)
stopifnot(abs(calc_pct - found_pct) < 1e-8)
```

### 3.4 Airport Mix by Airline (Normalized stacked bar)

```{r}
#| label: airport-mix
mix <- departures %>%
  filter(
    airline %in% (departures_by_airline %>% filter(n >= 1000) %>% pull(airline))
  ) %>%
  group_by(airline, airport) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(airline) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

mix <- mix %>%
  mutate(
    airline = factor(
      airline,
      levels = (departures_by_airline %>%
        arrange(on_time_pct) %>%
        pull(airline))
    )
  )

ggplot(mix, aes(x = pct, y = fct_rev(airline), fill = airport)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = ifelse(pct >= 5, sprintf("%.0f%%", pct), "")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  labs(
    x = "Percent of departures",
    y = "Airline",
    title = "Airport mix by airline (normalized to 100%)"
  ) +
  theme_minimal()

# Verifier: percentages per airline sum to ~100
sum_check <- mix %>% group_by(airline) %>% summarise(total = sum(pct))
stopifnot(all(
  abs(sum_check$total - 100) < 1e-6 | abs(sum_check$total - 100) < 0.1
))
```

### 3.5 On-Time Departures by Airport

```{r}
#| label: on-time-by-airport
by_airport <- departures %>%
  group_by(airport) %>%
  summarise(n = n(), on_time_pct = mean(on_time) * 100, .groups = "drop")

ggplot(
  by_airport,
  aes(x = reorder(airport, -on_time_pct), y = on_time_pct, fill = airport)
) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%", on_time_pct)), vjust = -0.5) +
  labs(
    x = "Airport",
    y = "On-time departure percentage",
    title = "On-time departures by airport"
  ) +
  theme_minimal()
```

### 3.6 On-Time Departures by Time of Day and Airport (Heatmap)

```{r}
#| label: heatmap-hour-airport
# compute on-time percentage by airport and hour
hour_airport <- departures %>%
  group_by(airport, hour) %>%
  summarise(n = n(), on_time_pct = mean(on_time) * 100, .groups = "drop")

ggplot(hour_airport, aes(x = hour, y = airport, fill = on_time_pct)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C") +
  labs(
    x = "Scheduled hour",
    y = "Airport",
    fill = "On-time %",
    title = "On-time % by scheduled hour and airport"
  ) +
  theme_minimal()
```

### 3.7 Schedule Mix by Airline (Hourly profile)

```{r}
#| label: schedule-mix
# hourly schedule mix (percent) for airlines with >=1000 departures
hourly_mix <- departures %>%
  filter(
    airline %in% (departures_by_airline %>% filter(n >= 1000) %>% pull(airline))
  ) %>%
  group_by(airline, hour) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(airline) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(hourly_mix, aes(x = hour, y = pct)) +
  geom_line() +
  facet_wrap(~airline, ncol = 3) +
  labs(
    x = "Scheduled hour",
    y = "Percent of airline's departures",
    title = "Schedule mix by hour for major airlines"
  ) +
  theme_minimal()
```

### 3.8 Observed versus Expected On-Time Departure Percentage

```{r}
#| label: observed-vs-expected
# overall on-time rates by airport and hour (pooled)
base_rates <- departures %>%
  group_by(airport, hour) %>%
  summarise(base_on_time = mean(on_time), .groups = "drop")

# airline exposure to each airport-hour
airline_exposure <- departures %>%
  group_by(airline, airport, hour) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(airline) %>%
  mutate(weight = n / sum(n)) %>%
  ungroup()

# expected on-time pct for each airline
expected <- airline_exposure %>%
  left_join(base_rates, by = c("airport", "hour")) %>%
  group_by(airline) %>%
  summarise(
    expected_on_time = sum(weight * base_on_time) * 100,
    .groups = "drop"
  )

observed <- departures %>%
  group_by(airline) %>%
  summarise(observed_on_time = mean(on_time) * 100, .groups = "drop")

obs_exp <- observed %>%
  left_join(expected, by = "airline") %>%
  left_join(departures_by_airline %>% select(airline, n), by = "airline")

ggplot(
  obs_exp %>% filter(n >= 1000),
  aes(x = expected_on_time, y = observed_on_time, label = airline)
) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_text_repel() +
  labs(
    x = "Expected on-time % (given airport-hour mix)",
    y = "Observed on-time %",
    title = "Observed vs Expected On-time % by Airline"
  ) +
  theme_minimal()

# verifier: ensure obs_exp rows match number of airlines
stopifnot(nrow(obs_exp) == nrow(departures_by_airline))
```
